WSClean: an implementation of a fast, generic wide-field imager

### 2014-03-14 articles/2014/Article-WSClean/wsclean.tex
Steven's comments

\section{Introduction}
Visibility data from non-coplanar interferometric radio telescopes which observe large fractions of the sky at once can not be accurately imaged with a two-dimensional fast Fourier transform (FFT). Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. The image degradation effects of the $w$-term are amplified for telescopes with wide field of view (FOV), making this a particularly important issue for low-frequency telescopes that by nature are wide-field instruments.

There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, $w$-projection \citep{wprojection-cornwell}, $w$-stacking \citep{ska-memo-regridding-2011} and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (\textsc{casa}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, with the MWA we have seen that imaging a 2-min snapshot observation away from zenith can take up to tens of hours with \textsc{casa}'s $w$-projection algorithm, because of the larger $w$-terms for off-zenith observations. Imaging with larger image sizes or at higher zenith angles can be impossible because the size and number of $w$-kernels become too large to hold in memory.

@@

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for wide-field telescopes. One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
One particular DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}. This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \textsc{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking \citep{awimager-2013}.

@@

To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~applying the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeating this for all $w$-values and adding the results together; v)~applying the final scaling.

@@

\subsection{Discretization of $w$} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. \textsc{wsclean} was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of $w$-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal. This is in contrast to \citet{wprojection-cornwell} where $\sqrt{w}$ tabulation is suggested. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlay} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can be very different, depending on the observation. The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle (ZA) and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, and reaches $\sim 400$ at a ZA of $30^{\circ}$. For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75'$ size, the $\max_{l,m}$ term is $0.68$. This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging FOV.

\begin{table}
 \caption{Scaling of the computational cost for various imaging steps} \label{tbl:computational-cost-per-operation}
 \begin{tabular}{lll}
   \textbf{Operation} & \textbf{$w$-stacking} & \textbf{$w$-projection} \\
   \hline\hline
   Fourier transform(s) & $N_\textrm{wlay} N^2_\textrm{pix} \log N_\textrm{pix}$ & $N^2_\textrm{pix} \log N_\textrm{pix}$ \\
   $w$-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
   Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$ \\
   \hline\hline
 \end{tabular}
\end{table}

@@

To efficiently image the full sky, an observation is split into short snapshots, each of which is phase-rotated to zenith and subsequently imaged, with an appropriate number of pixels and resolution. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10k pixels along each side. The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value. For the 128-tile MWA, the maximum differential elevation between tiles is $8.5$m. The tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$. In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.

@@

An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with \textsc{wsclean}. The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately. Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy. The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image. Of the 41 min, 3 min is spent on phase rotating the visibilities.

@@

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. \textsc{casa} convolves samples with a PSWF of seven pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate. \textsc{wsclean} currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}. After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that seven pixels is a good compromise between performance and accuracy, and this is the default for \textsc{wsclean}.

\subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
Typically, for cleaning it is desired that images have a pixel size at least five times smaller than the synthesized-beam width. This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions. This factor of five is normally taken into account in the overall imaging resolution, i.e., the image size is increased during inversion. For the $w$-projection method, the cost of inversion with an increased resolution is small, because it does not increase the size of the $w$-kernels. It will affect the inverse FFT, but the relative cost of this step is negligible in the total cost. The $w$-stacking method is however significantly affected by the image resolution, because it performs many inverse FFTs.

The spatial frequencies in the output image are band-limited by the synthesized beam. Therefore, it is not required to perform the inversion at higher resolution. As long as the $uv$-plane is sampled with at least the Nyquist frequency, a high resolution image can be perfectly reconstructed from an inversion at lower resolution. After an enlarged image has been cleaned, a model is created at the same (high) resolution. Because the model consists of delta functions, the spatial frequencies in the model are not band-limited. Consequently, shrinking the model image will remove information from the model. However, only the low spatial frequencies of this image will be used in the prediction. Therefore, inversion at higher resolution is not required: as long as long as downscaling does not modify the low-frequency components, the output of the prediction step will not change. This requires low-pass filtering the model before decimation.

We have implemented an option in \textsc{wsclean} to automatically decrease the inversion resolution to the Nyquist limit,
\begin{equation}\label{eq:nyquist-resolution}
 N_\textrm{Nyquist} = 2 \frac{N_\textrm{pix} S_\textrm{pix}}{S_\textrm{synth beam}},
\end{equation}
where $N_\textrm{Nyquist}$ the resolution in pixels used during inversion, $N_\textrm{pix}$ the requested number of pixels of the image along one side, $S_\textrm{pix}$ the requested angular pixel scale and $S_\textrm{synth beam}$ the minimum angular size of the synthesized beam. Before cleaning, the low-resolution image is interpolated with a procedure consisting of i) a low-resolution FFT; ii) zero padding; and iii) a high-resolution inverse FFT. Such interpolation assumes the image to be periodic, but this is already assumed during the inversion process. After cleaning, the model is decimated by the inverse procedure, consisting of i) a high-resolution FFT; ii) truncation; and iii) a low-resolution inverse FFT. With common settings, this procedure decreases the inversion imaging resolution with a factor of 2.5. Without further correction, such a decrease would lower the positional accuracy with which visibilities are gridded onto the $uv$-plane. This can be corrected by increasing the oversampling rate, which has almost no effect on performance.

Recreating Fig.~\ref{fig:vela-projection-example} using both the snapshot method and this optimization lowers the computational cost from 41~min to 24~min. Of the 24~min, 77\% is spent on cleaning approximately 100,000 components. The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy, while the total RMS is 227 mJy. We think this can be explained by the non-linear behaviour of clean.

\section{Analysis} \label{sec:analysis}
We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in \textsc{casa}. For the analysis, we use imaging parameters which are common for MWA imaging. When a specific parameter value is not mentioned, the settings from Table~\ref{tbl:default-parameters} are used. The number of $w$-projection planes in $w$-projection is kept equal to the number of $w$-layers in $w$-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}. This yields 195 $w$-planes/layers at 10\degree ZA. Several configurations with large $w$-values fail to image with \textsc{casa}, presumably because the $w$-kernels become too large.

The used software version of \textsc{casa} is ``stable release 42.0, revision 26465'', which was released September 2013. For \textsc{wsclean}, the latest version as of February 2014 was used. The tests were run on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores that can perform 138 giga-floating point operations per second (GFLOPS). The data are stored on a multi-disk array with five spinning hard disks, which has a combined read rate of about 450 MB/s.
\begin{table}%
\caption{Parameter values used during benchmarks, unless otherwise mentioned.} \label{tbl:default-parameters}%
\begin{center}\begin{tabular}{rl}%
\hline
Array & MWA \\
Number of elements & 128 \\
Image size & $3072 \times 3072$ \\
Angular pixel size & $0.72'$ \\
Number of visibilities & $3.5 \times 10^8$ \\
Time resolution & 2~s \\
Frequency resolution & 40~kHz \\
Observation duration & 112 s\\
Bandwidth & 30.72 MHz (768 channels)\\
Central frequency & 182 MHz \\
Zenith angle at phase centre & 10\degree \\
Max $w$-value for phase centre & 172 $\lambda$ (283 m) \\
Number of polarizations in set & 4 \\
Imaged polarization & $pp$ ($\sim$XX) \\
Imaging mode & multi-frequency synthesis \\
Weighting & uniform \\
Data size & 18 GB \\
\hline
\end{tabular}\end{center}\end{table}

@@

Table~\ref{tbl:accuracy-measurements} lists the measured root mean square (RMS) in the residual image and the standard errors of the source brightnesses as detected by \textsc{Aegean}. In all tests, \textsc{wsclean} performs slightly more accurately. At zenith, \textsc{casa} produces a residual RMS twice as large as \textsc{wsclean}. This is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith. This evidently has a stronger effect on the $w$-projection algorithm. However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more $w$-projection planes. When 128 $w$-projection planes are used in \textsc{casa}, the residual RMS is equal to \textsc{wsclean} with 12 $w$-layers, but the flux density measurements are less accurate. We do not know why this parameter needs to be higher in \textsc{casa} to reach the same RMS. We are using enough $w$-planes to cover the sources: Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlay}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree. Also unexpected is that the source flux density becomes worse by increasing the number of planes. For \textsc{wsclean} both values stay approximately the same when the number of $w$-layers is increased.

>>>>

\section{Introduction}
Visibility data from non-coplanar interferometric radio telescopes which observe large fractions of the sky at once can not be accurately imaged with a two-dimensional fast Fourier transform (FFT). Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. The image degradation effects of the $w$-term are amplified for telescopes with wide fields of view (FOV), making this a significant issue for low-frequency telescopes that by nature are wide-field instruments.

There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}; a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}; $w$-projection \citep{wprojection-cornwell}; $w$-stacking \citep{ska-memo-regridding-2011}; and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.

A new generation of wide-field observatories is producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (\textsc{casa}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, with the MWA we have seen that imaging a 2-min snapshot observation away from zenith can take up to tens of wall-clock hours with \textsc{casa}'s $w$-projection algorithm, because of the larger $w$-terms for off-zenith observations. Imaging with larger image sizes or at higher zenith angles can be impossible because the size and number of $w$-kernels become too large to hold in memory.

@@

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for wide-field telescopes. One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
One particular DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all antennas, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}. This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \textsc{awimager} \citep{awimager-2013} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking.

@@

To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by: i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~applying the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeating this for all $w$-values and adding the results together; v)~applying the final scaling.

@@

\subsection{Discretization of $w$} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. \textsc{wsclean} was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of $w$-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggests that a uniform discretization in $w$ is optimal. This is in contrast to \citet{wprojection-cornwell} where $\sqrt{w}$ tabulation is suggested. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlay} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can be very different, depending on the observation. The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle (ZA) and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, and reaches $\sim 400$ at a ZA of $30^{\circ}$. For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75'$ size, the $\max_{l,m}$ term is $0.68$. This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
\begin{table}
 \caption{Scaling of the computational cost for various imaging steps, with $N_\textrm{wlay}$ the number of $w$-layers, $N_\textrm{pix}$ the number of pixels along each side, $N_\textrm{vis}$ the number of visibilities, $N_\textrm{kern}$ the size of the anti-aliasing kernel, $N_\textrm{wkern}$ the size of the $w$-kernel, $w_{\max}$ the maximum $w$ value and $\alpha_\textrm{FOV}$ the imaging FOV.} \label{tbl:computational-cost-per-operation}
 \begin{tabular}{lll}
   \textbf{Operation} & \textbf{$w$-stacking} & \textbf{$w$-projection} \\
   \hline\hline
   Fourier transform(s) & $N_\textrm{wlay} N^2_\textrm{pix} \log N_\textrm{pix}$ & $N^2_\textrm{pix} \log N_\textrm{pix}$ \\
   $w$-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
   Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$ \\
   \hline\hline
 \end{tabular}
\end{table}

It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging FOV.

@@

To efficiently image the full sky, an observation is split into short snapshots, their phase centres are changed to zenith and the snapshots are subsequently imaged, with an appropriate number of pixels and resolution. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10k pixels along each side. The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value. For the 128-tile MWA, the maximum differential elevation between tiles is $8.5$m. The tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$. In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.

@@

An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with \textsc{wsclean}. The processing steps performed for this image were: i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately. Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy. The wall-clock time for imaging is 60 min and 41 min for normal projection and the recentred image, respectively. Of the 41 min, 3 min is spent on phase rotating the visibilities.

@@

The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}. \textsc{casa} convolves samples with a PSWF of seven pixels total width during gridding. When using a variable kernel size, a PSWF is quite complicated and computationally expensive to calculate. \textsc{wsclean} currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}. After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that seven pixels is a good compromise between performance and accuracy, and this is the default for \textsc{wsclean}.

\subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
Typically, for cleaning it is desired that images have a pixel size (side length) at least five times smaller than the synthesized-beam width. This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions. This factor of five is normally taken into account in the overall imaging resolution, i.e., the image size is increased during inversion. For the $w$-projection method, the cost of inversion with an increased resolution is small, because it does not increase the size of the $w$-kernels. It will affect the inverse FFT, but the relative cost of this step is negligible in the total cost. The $w$-stacking method is however significantly affected by the image resolution, because it performs many inverse FFTs.

The spatial frequencies in the output image are band-limited by the synthesized beam. Therefore, it is not required to perform the inversion at higher resolution. As long as the $uv$-plane is sampled with at least the Nyquist frequency, a high resolution image can be perfectly reconstructed from an inversion at lower resolution. After an enlarged image has been cleaned, a model is created at the same (high) resolution. Because the model consists of delta functions, the spatial frequencies in the model are not band-limited. Consequently, decimating the model image will remove information from the model. However, only the low spatial frequencies of this image will be used in the prediction. Therefore, inversion at higher resolution is not required: as long as the decimation does not modify the low-frequency components, the output of the prediction step will not change. This requires low-pass filtering of the model before decimation.

We have implemented an option in \textsc{wsclean} to automatically decrease the inversion resolution to the Nyquist limit,
\begin{equation}\label{eq:nyquist-resolution}
 N_\textrm{Nyquist} = 2 \frac{N_\textrm{pix} S_\textrm{pix}}{S_\textrm{synth beam}},
\end{equation}
where $N_\textrm{Nyquist}$ is the resolution in pixels used during inversion, $N_\textrm{pix}$ is the requested number of pixels of the image along one side, $S_\textrm{pix}$ is the requested angular pixel scale and $S_\textrm{synth beam}$ is the minimum angular size of the synthesized beam. Before cleaning, the low-resolution image is interpolated with a procedure consisting of: i) a low-resolution FFT; ii) zero padding; and iii) a high-resolution inverse FFT. Such interpolation assumes the image to be periodic, but this is already assumed during the inversion process. After cleaning, the model is decimated by the inverse procedure, consisting of: i) a high-resolution FFT; ii) truncation; and iii) a low-resolution inverse FFT. With common settings, this procedure decreases the inversion imaging resolution by a factor of 2.5. Without further correction, such a decrease would lower the positional accuracy with which visibilities are gridded onto the $uv$-plane. This can be corrected by increasing the oversampling rate, which has almost no effect on performance.

Recreating Fig.~\ref{fig:vela-projection-example} using both the snapshot method and this optimization lowers the computational cost from 41~min to 24~min. Of the 24~min, 77\% is spent on cleaning approximately 100,000 components. The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy, while the total RMS is 227 mJy. Because the difference is mostly noise like, the difference could be caused by the non-linear behaviour of clean.

\section{Analysis} \label{sec:analysis}
We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in \textsc{casa}. For the analysis, we use imaging parameters common for MWA imaging. When a specific parameter value is not mentioned, the settings from Table~\ref{tbl:default-parameters} are used. The number of $w$-projection planes in $w$-projection is kept equal to the number of $w$-layers in $w$-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}. This yields 195 $w$-planes/layers at 10\degree ZA. Several configurations with large $w$-values fail to image with \textsc{casa}, because \textsc{casa} crashes during the imaging, presumably because the $w$-kernels become too large.

The software version of \textsc{casa} is ``stable release 42.0, revision 26465'', which was released September 2013. For \textsc{wsclean}, version 1.0 from February 2014 was used. The tests were run on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores that can perform 138 giga-floating point operations per second (GFLOPS). The data are stored on a multi-disk array with five spinning hard disks, which has a combined read rate of about 450 MB/s.
\begin{table}%
\caption{Parameter values used during benchmarks, unless otherwise mentioned.} \label{tbl:default-parameters}%
\begin{center}\begin{tabular}{rl}%
\hline
Array & MWA \\
Number of elements & 128 \\
Image size & $3072 \times 3072$ \\
Angular pixel size & $0.72'$ \\
Number of visibilities & $3.5 \times 10^8$ \\
Time resolution & 2~s \\
Frequency resolution & 40~kHz \\
Observation duration & 112 s\\
Bandwidth & 30.72 MHz (768 channels)\\
Central frequency & 182 MHz \\
Zenith angle at phase centre & 10\degree \\
Max $w$-value for phase centre & 172 $\lambda$ (283 m) \\
Number of polarizations in set & 4 \\
Imaged polarization & $pp$ ($\sim$XX) \\
Imaging mode & multi-frequency synthesis \\
Weighting & uniform \\
Data size & 18 GB \\
\hline
\end{tabular}\end{center}\end{table}

@@

Table~\ref{tbl:accuracy-measurements} lists the measured root mean square (RMS) in the residual image and the standard errors of the source brightnesses as detected by \textsc{Aegean}. \textsc{wsclean} is more accurate: it shows 2-33\% lower errors in the source fluxes and produces 0-49\% lower RMS noise compared to \textsc{casa}. The large residual RMS twice for \textsc{casa} at zenith is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith. This evidently has a stronger effect on the $w$-projection algorithm. However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more $w$-projection planes. When 128 $w$-projection planes are used in \textsc{casa}, the residual RMS is equal to \textsc{wsclean} with 12 $w$-layers, but the flux density measurements are less accurate. We do not know why this parameter needs to be higher in \textsc{casa} to reach the same RMS. We are using enough $w$-planes to cover the sources: Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlay}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree. Also unexpected is that the source flux density becomes worse by increasing the number of planes. For \textsc{wsclean} both values stay approximately the same when the number of $w$-layers is increased.

In the $\textrm{ZA}=10\degree$ case, \textsc{casa} is slightly less accurate. Extra noise can be seen in the images, as shown in Fig.~\ref{fig:residuals}. An image resulting from the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analysed. As can be seen in Table~\ref{tbl:accuracy-measurements}, the source fluxes in the recentred image have smaller errors compared to the normal projection, but the residual noise is higher. The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy. We have performed tests with and without the optimization of \S\ref{sec:decreasing-inversion-resolution}. They yield identical numbers.

\textsc{wsclean} is faster in all tested cases. Both imagers perform five major iterations for these results, and the inversions and predictions are dominating the computing time. We will look more closely at the differences in performance in the next section.
